%\subsection* {Constructor} %------------------------------------
%\subsection* {Methods to Override} %----------------------------
%\subsection* {Utility Methods} %--------------------------------
%\subsection* {Other Methods} %----------------------------------
%\subsection* {Inherited Methods} %------------------------------
%
%
\documentclass[letterpaper,draft]{book}
% Try to fill up the page!
% This was from another example - we will ignore it for now
\normalsize%\setlength{\voffset}{-.5in}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{1in}
\setlength{\evensidemargin}{.05in}
\setlength{\marginparwidth}{0in}
\addtolength{\textheight}{1in}
\addtolength{\textwidth}{.7in}
% \code is a piece of code
\newcommand{\code}{\texttt}
% \V is V
\newcommand{\V}{\emph{\textbf{V}}}
% special math chars I need: -> ~ ^
\newcommand{\rta}{$\rightarrow$\ }
\newcommand{\tild}{\~{ }}
%\newcommand{\vrbsize}{\small}
\newcommand{\verbfont}{\footnotesize}
\newcommand{\regfont}{\normalfont\normalsize}
% \Class is used to start a new class description at page begin

%----old \newcommand{\Class}[1]{\newpage \markboth{#1}{#1} \section* {#1}}
\newcommand{\Class}[1]{\section {#1}}

%\Indextt for tt index

\newcommand{\Indextt}[1]{\index{#1@\texttt{#1}}}
% \Meth is for a method
\newcommand{\Meth}[1]{\subsubsection* {#1}}

% \Cmd is for a dialog command
\newcommand{\Cmd}[1]{\subsubsection* {#1}}

% \Param is a parameter description
\newcommand{\Param}[1]{\subparagraph {#1}}

% kill off the figfont
\newcommand{\SetFigFont}[3]{}%

% \Sect is used to set typeface of chapter and section titles
\newcommand{\Sect}[1]{\emph{#1}}

% Method environment that allows parameters to be entered with
% the \Param command between the \begin and \end.
\newenvironment{Method}[1]{\Meth{#1}
\begin{description}
\begin{description}}
{\end{description}
\end{description}}

% An environment for a definition
\newenvironment{Def}[1]{\begin{description} \item[#1]}{\end{description}}

\newenvironment{Point}{\begin{itemize} \item }{\end{itemize}}

%\usepackage{verbatim}
%\usepackage{eepic}
%\usepackage{pictex}
\usepackage{makeidx}

\input{html.sty}
%--------
\makeatletter
\def\@evenhead{
	\textbf{\textrm{\thepage \hfill \leftmark}}}
\def\@oddhead{
	\textbf{\textrm{\rightmark \hfill \thepage}}}
\makeatother

\renewcommand{\chaptermark}[1]{\markboth{#1}{#1}}
\renewcommand{\sectionmark}[1]{\markright{\thesection~#1}}

\newfont{\REALLYHUGE}{cminch}
\makeindex
%==========================================================================
\begin{document}

\vspace*{1in} \thispagestyle{empty}
\begin{center}

\REALLYHUGE V \Huge \\ ~ \\ \textbf{A C++ GUI \\ Framework}

\vspace{1in}

\LARGE Bruce E. Wampler, Ph.D. \\
\large ~ \\
 \emph{P.O. Box 140} \\ 
 \emph{35 High Country Dr.} \\ 
 \emph{Cedar Crest, NM 87008} \\ ~ \\
 \texttt{wampler@cs.unm.edu} \\
 \texttt{http://www.cs.unm.edu/\~{}wampler}

\vspace{1in}
\normalsize
Version 1.16 \\
January 20, 1997
\end{center}

\newpage
\vspace*{1in} \thispagestyle{empty}

\textbf{V - A C++ GUI Framework}

Copyright \copyright 1995,1996,1997, Bruce E. Wampler

All rights reserved.

\texttt{wampler@cs.unm.edu}

\texttt{http://www.cs.unm.edu/\~{}wampler}

\vspace{1in}

\emph{V - A C++ GUI Framework}, Version 1.16,
may be reproduced and distributed, in whole or in part, subject
to the following conditions:

\begin{enumerate}
\item The copyright notice above and this permission notice must
be preserved complete on all complete or partial copies.

\item You may not translate or create a derivative of this work
without the author's written permission.

\item If you distribute this manual in part, you must provide
instructions and a means for obtaining a complete version.

\item You may make a profit on copies of this work only if it
is included as part of an electronic distribution of other free software
works (e.g., Linux or GNU).

\item Small portions may be reproduced as illustrations for reviews
or quotations in other works without this permission notice if proper
citation is given.

\end{enumerate}

My goal is to get as many people as can be helped using \V\@. If
the terms of this documentation copyright are unsatisfactory, please
contact me and we can probably work something out.

The complete source for \V\ (X Athena and Microsoft Windows versions)
is available via anonymous ftp at \code{ftp.cs.unm.edu/pub/wampler}, or
via my www home page (see above). This manual
is also available as an html document on my web page.

I would like to acknowledge the Computer Science Department of the
University of New Mexico for providing e-mail, ftp, and WWW service.

\pagenumbering{roman} \setcounter{page}{2} \tableofcontents

\input{vrefch1.tex}
\input{vrefch2.tex}
\input{vrefch3.tex}
\input{vrefch4.tex}
\input{vrefch5.tex}
\input{vrefch6.tex}
\input{vrefch7.tex}
\input{vrefch8.tex}
\input{vrefch9.tex}

%***********************************************************************
%***********************************************************************
%***********************************************************************

\appendix


\chapter{Tutorial C++ Source}

\input{tutapndx.tex}

\input{cxxstyle.tex}

\input{quickref.tex}

\input{vtools.tex}

\input{release.tex}
%***********************************************************************
%***********************************************************************
%***********************************************************************
\chapter{V Class Hierarchy}

The following figure is shows the internal organization of the \V\
class hierarchy. Note that boxes with a double line edge denote
classes that
have object instances, while boxes with single line edges are abstract
classes used to build subclasses.  Most of the time, you won't care
about these abstract classes.

Also note that the classes derived from \code{vCmd} represent the
classes used to implement command objects. Normally, you won't need
to use these classes directly. Instances of these objects are generated
by \V\ from your \code{CommandObject} declarations when you call
\code{AddDialogCmds}.
\newpage
\begin{center}
{\Large The V GUI Class Hierarchy}
\end{center}

\begin{center}
\vspace{.1in}
\small
\input{fig/vapphier.tex}
\regfont
\end{center}
\printindex
\end{document}
