\documentstyle[12pt]{article}
\title{\bf Xspread}
\author{Rama Devi Puvvada}
\date{10 August 1994}

\begin{document}
\maketitle
Xspread is a public domain spreadsheet program on the X window system similar 
to Lotus 1-2-3. At present Xspread is already useable and has no major defects 
except for a few bugs/deficiencies. My project as a part of the course work of 
the Software Engineering course, summer '94 is to make some improvements and 
enhancements to the existing software.
\\
\\
The following are some of the improvements I made.
    \begin{itemize}
     \item Ran lint on all the c programs to remove unreasonable constructs.
     \item Ability to change the foreground and the background colors of the spreadsheet window.
     \item Corrected the errors in plotting graphs when the size of the ranges do not match.
     \item Added colors to the graphs.
     \item Improved the efficiency of the matrix functions. 
     \item Sort the table by rows or by columns in increasing or decreasing order and using label or value.
     \item Search a label or value in the entire table or in a specific range in Row major or Column major order .
    \end{itemize}.
\\
\\
The demonstration includes the following steps.
\begin{itemize}
  \item START
   \begin{itemize}
    \item Start the spreadsheet with -fg and -bg options (foreground and background options respectively).
    \item Go to main menu.
    \item Select file menu and input file 'ex1'.
   \end{itemize}
  \item COLOR
   \begin{itemize}
    \item Go to main menu.
    \item Select Color.
    \item Give the foreground and background colors and press ESC key.
    \item Map the window.
   \end{itemize}
  \item GRAPH
   \begin{itemize}
    \item Go to main menu.
    \item Go to Graph menu and view the Gragh (I already set all the options like setting the ranges etc.. to draw a Stack Bar Graph in the file 'ex1' to save time).
    \item Go to Type and change it to Bar graph and view it.
    \item Go to Type and change it pie graph and view it.
    \item Go to File menu and input file 'ex2' and view the graph. It has options set for a line graph.
    \item Go to File menu and input file 'ex3' and view the graph. It has options set for a XY graph.
   \end{itemize}
  \item SORT
   \begin{itemize}
    \item Go to main menu.
    \item Select the Sort menu.
    \item By default the options will be set to increasing order and sort by value.
    \item Select the Column/Row option.
    \item Select the column and input the column number. 
    \item The given table is sorted by that column in increasing order and by value.
    \item To try for other types, Go to the Sort menu, select decreasing order, sort by label, and by row and input the row number.
    \item The table is sorted with the given options.
   \end{itemize}
  \item SEARCH
   \begin{itemize}
    \item Go to main menu.
    \item Select the Search menu.
    \item By default the options will be set to Entire table and in Column major order. 
    \item Enter the value and start the search by selecting Next or Previous.
    \item Change the options to Row major and give a specific Range to search for a label and continue the search.
   \end{itemize}
  \end{itemize}
\end{document}

