%$Id: w.tex,v 1.1 96/03/24 11:12:43 al Exp $
% man circuit w .
%------------------------------------------------------------------------
\section{{\tt W}: Current Controlled Switch}
%------------------------------------------------------------------------
\subsection{Syntax}
\begin{verse}
{\tt W}{\it xxxxxxx n+ n-- ce mname} \{{\it ic}\}
\end{verse}
%------------------------------------------------------------------------
\subsection{Purpose}

Current controlled switch.
%------------------------------------------------------------------------
\subsection{Comments}

{\it N+} and {\it n--} are the positive and negative element nodes,
respectively.  {\it Ce} is the name of an element through which
the controlling current flows.  {\it Mname} is the model name.  A
switch is a resistor between {\it n+} and {\it n--}.  The value of
the resistor is determined by the state of the switch.

The resistance between {\it n+} and {\it n--} will be {\it RON}
when the controlling current (through {\it ce}) is above {\it IT}
+ {\it IH}.  The resistance will be {\it ROFF} when the controlling
current is below {\it IT} - {\it IH}.  When the controlling current
is between {\it IT} - {\it IH} and {\it IT} + {\it IH}, the resistance
will retain its prior value.

You may specify {\tt ON} or {\tt OFF} to indicate the initial state
of the switch when the controlling current is in the hysteresis
region.

{\tt RON} and {\tt ROFF} must have finite positive values.

The controlling element can be any simple two terminal element.
Unlike SPICE, it does not need to be a voltage source.
%------------------------------------------------------------------------
\subsection{Model Parameters}

\begin{description}

\item[{\tt IT} = {\it x}] Threshold current.  (Default = 0.)

\item[{\tt IH} = {\it x}] Hysteresis current.  (Default = 0.)

\item[{\tt RON} = {\it x}] On resistance.  (Default = 1.)

\item[{\tt ROFF} = {\it x}] Off resistance.  (Default = 1e12)

\end{description}
%------------------------------------------------------------------------
%------------------------------------------------------------------------
